---
title: "data cleaning"
format: html
---

## setup

Load required packages:

```{r}
library(readr)
library(tidyverse)
library(here)
library(rvest)
library(readxl)
library(janitor)
library(scales)
library(haven)
```

## reading in and tidying

fips data for merging

```{r}
fips <- tibble(tidycensus::fips_codes) |> 
  select(state_code, state, state_name) |> 
  unique()
```

Read in house and senate gerrymandering data from PlanScore.org

```{r}
bias_statehouse <- read_csv("https://planscore.org/data/bias_statehouse.csv")
bias_statesenate <- read_csv("https://planscore.org/data/bias_statesenate.csv")
```

get bias data into tidy stateyear format

```{r}
bias_h <- bias_statehouse |> 
  select(state, year, code, plan, eg_adj_avg) |> 
  rename(eg_h = eg_adj_avg,
         plan_h = plan,
         code_h = code) |>
  arrange(state,year) |> 
  group_by(state) |> 
  mutate(eg_h = -1*eg_h, # republican = positive
         planchange = if_else(lead(plan_h) == plan_h, 0, 1),
         reps = case_when(planchange == 0 ~ lead(year) - year,
                          planchange == 1 ~ 2,
                          .default = 1)) |> 
  uncount(reps) |> 
  group_by(code_h) |> 
  mutate(year = year[1] + row_number() - 1) |> 
  select(-planchange)
  
bias_s <- bias_statesenate |> 
  select(state, year, code, plan, eg_adj_avg) |> 
  rename(eg_s = eg_adj_avg,
         plan_s = plan,
         code_s = code) |>
  arrange(state,year) |> 
  group_by(state) |> 
  mutate(eg_s = -1*eg_s, #republican = positive
         planchange = if_else(lead(plan_s) == plan_s, 0, 1),
         reps = case_when(planchange == 0 ~ lead(year) - year,
                          planchange == 1 ~ 2,
                          .default = 1)) |> 
  uncount(reps) |> 
  group_by(code_s) |> 
  mutate(year = year[1] + row_number() - 1) |> 
  select(-planchange)

bias <- full_join(bias_h, bias_s, by = c("state", "year")) |> 
  rowwise() |> 
  mutate(eg = mean(c(eg_h, eg_s), na.rm = F)) |> 
  ungroup() |> 
  mutate(eg_abs = abs(eg),
         eg_z = datawizard::standardize(eg),
         eg_abs_z = datawizard::standardize(eg_abs)) |> 
# adding planyear variable for merging with partisanship, then limiting to 1990+
  arrange(state, year) |> 
  group_by(state) |> 
  fill(plan_s, .direction = "down") |> 
  fill(plan_h, .direction = "down") |> 
  mutate(planyear = case_when(lag(plan_h) != plan_h ~ year,
                              lag(plan_s) != plan_s ~ year,
                              .default = NA_integer_)) |> 
  relocate(planyear, .after = "year") |> 
  fill(planyear, .direction = "down") |> 
  ungroup() |> 
  filter(year > 1989)
```

income inequality data

```{r}
ii <- read_delim(here("data", "WID_Data_16052024-192751.csv"), delim = ";", 
           skip = 1, col_names = FALSE) |> 
  select(X1, X4, X5) |> 
  rename(state_name = X1,
         year = X4,
         top10share = X5) |> 
  mutate(state_name = case_match(state_name,
                                 "Georgia (US)" ~ "Georgia",
                                 .default = state_name)) |> 
  left_join(fips) |> 
  select(state, year, top10share)
```

partisanship data

```{r}

# reading in the original data
partisan_1 <- read_xls(here("data", "partisan", "12s0419.xls"), 
         skip = 3, n_max = 54, col_names = FALSE) |> 
  mutate(across(everything(), ~str_remove_all(.x, "[[:space:]](\\\\)[[:digit:]]{1,}"))) |> 
  t() |> 
  as.data.frame() |> 
  tibble() |> 
  row_to_names(1) |> 
  clean_names(case = "title") |> 
  rename(house = State, 
         year = Na, 
         party = Na_2) |> 
  fill(house, .direction = "down") |> 
  fill(year, .direction = "down") |> 
  pivot_longer(!c(house,year,party),
               names_to = "state",
               values_to = "seats") |> 
  pivot_wider(names_from = party, 
              values_from = seats) |> 
  mutate(across(Democrat:Vacancies, as.numeric)) |> 
  # throws a warning here: it's just Nebraska
  rowwise() |> 
  mutate(prop_rep = Republican / (Democrat + Republican + Other)) |> 
  group_by(state, year) |> 
  summarize(prop_rep = mean(prop_rep)) |>
  mutate(year = as.numeric(year),
         reps = lead(year) - year,
         reps = case_when(is.na(reps) ~ 1,
                          .default = reps)) |> 
  uncount(reps) |> 
  group_by(state, year) |> 
  mutate(year = year[1] + row_number() - 1)

# i have data from 1990-2011; anything after that, i'm going to have to pull individually from the NCSL for that year

# pulled archive PDFs from https://www.ncsl.org/about-state-legislatures/state-partisan-composition

# used pdf24 to convert to excel

read_partisanship <- function(fileyear){
read_xlsx(here("data", "partisan", paste0("LegisControl_", fileyear, ".xlsx")), 
          skip = 2, n_max = 50) |> 
  rename(state = 1,
         senatedem = 4,
         senaterep = 5,
         senateother = 6,
         housedem = 8,
         houserep = 9,
         houseother = 10) |> 
  select(c(state, senatedem, senaterep, senateother, 
           housedem, houserep, houseother)) |> 
  mutate(across(c(senateother,houseother), ~str_remove_all(.x, "[1:9]*v")),
         across(c(senateother,houseother), ~str_remove_all(.x, ", ")),
         across(senatedem:houseother, as.numeric),
         across(c(houseother, senateother), ~case_match(.x,
                                                        NA_integer_ ~ 0,
                                                        .default = .x))) |> 
  rowwise() |> 
  mutate(prop_rep = (senaterep / sum(senatedem, senaterep, senateother) + 
                       houserep / sum(housedem, houserep, houseother)) / 2) |> 
  select(state, prop_rep) |> 
  mutate(year = fileyear)
}
  
partisan_2 <- map(2012:2019, read_partisanship) |> list_rbind() |> 
  mutate(state = str_remove(state, "\\*"))

partisan <- bind_rows(partisan_1, partisan_2) |> 
  left_join(fips, by = c("state" = "state_name")) |> 
  ungroup() |> 
  select(state = state.y, year, prop_rep)
```

state median income data

```{r}
inc <- read_xls(here("data", "medianhhinc.xls"), sheet = 2) |> 
  rename_with(~str_extract(.x, "(?<=MEHOINUS)[[:alpha:]]{2}"),
              !DATE) |> 
  pivot_longer(cols = !DATE, 
               names_to = "state",
               values_to = "inc") |> 
  mutate(year = year(DATE)) |> 
  select(-DATE)
```

# merging

```{r}
preds_full <- bias |>
  mutate(lag1year = year + 1,
         lag2year = year + 2,
         lag3year = year + 3) |> 
  left_join(partisan, by = c("state" = "state", "planyear" = "year")) |> 
  rename(prop_rep_planwriters = prop_rep) |> 
  mutate(prop_rep_planwriters_z = datawizard::standardize(prop_rep_planwriters))

preds <- preds_full |> 
  select(state, year, planyear, 
         lag1year, lag2year, lag3year,
         eg, eg_abs, eg_z, eg_abs_z,
         prop_rep_planwriters, prop_rep_planwriters_z)

covs <- partisan |> 
  left_join(ii) |> 
  left_join(inc) |> 
  mutate(across(prop_rep:inc, ~datawizard::standardize(.x), .names = "{.col}_z"))


save(preds, covs, file = here("data", "preds_covs.Rdata"))
```

reading in dependent variables

```{r}

im1 <- read_delim(here("data", "im", "Linked Birth  Infant Death Records, 1995-1998.txt")) |> 
  select(state = 2,
         state_code = 3,
         year = 4,
         race = 6,
         deaths = 8,
         births = 9) |> 
  mutate(across(c(year, deaths, births), ~as.numeric(.x)))

im2 <- read_delim(here("data", "im", "Linked Birth  Infant Death Records, 1999-2002.txt")) |> 
  select(state = 2,
         state_code = 3,
         year = 4,
         race = 6,
         deaths = 8,
         births = 9) |> 
  mutate(across(c(year, deaths, births), ~as.numeric(.x)))

im3 <- read_delim(here("data", "im", "Linked Birth  Infant Death Records, 2003-2006.txt")) |> 
  select(state = 2,
         state_code = 3,
         year = 4,
         race = 6,
         deaths = 8,
         births = 9) |> 
  mutate(across(c(year, deaths, births), ~as.numeric(.x)))

im4 <- read_delim(here("data", "im", "Linked Birth  Infant Death Records, 2007-2021.txt")) |> 
  select(state = 2,
         state_code = 3,
         year = 4,
         race = 6,
         deaths = 8,
         births = 9) |> 
  mutate(across(c(year, deaths, births), ~as.numeric(.x)))

im <- bind_rows(im1, im2, im3, im4)

outcome_a <- im |> 
  mutate(race = case_match(race,
                           "Non-Hispanic White" ~ "w",
                           "Non-Hispanic Black" ~ "b"),
         imr = 1000 * deaths / births) |> 
  select(-deaths, -births, -state_code) |> 
  drop_na(race) |> 
  pivot_wider(id_cols = c("state", "year"),
              names_from = "race",
              values_from = "imr",
              names_prefix = "imr_") |> 
  drop_na() |> 
  filter(year %in% c(1995:2018)) |> 
  arrange(state, year) |> 
  left_join(fips, by = c("state" = "state_name")) |> 
  select(state = state.y,
         year,
         imr_w,
         imr_b)
```

adding the total infant mortality rate

```{r}
imr_t_1 <- read_csv(here("data", "im", "imr_all_95to98.csv")) |> 
  select(state = 2,
         state_code = 3,
         year = 4, 
         deaths = 6,
         births = 7) |> 
  mutate(race = "t")

imr_t_2 <- read_csv(here("data", "im", "imr_all_99to02.csv")) |> 
  select(state = 2,
         state_code = 3,
         year = 4, 
         deaths = 6,
         births = 7) |> 
  mutate(race = "t")

imr_t_3 <- read_csv(here("data", "im", "imr_all_03to06.csv")) |> 
  select(state = 2,
         state_code = 3,
         year = 4, 
         deaths = 6,
         births = 7) |> 
  mutate(race = "t")

imr_t_4 <- read_csv(here("data", "im", "imr_all_07to23.csv")) |> 
  select(state = 2,
         state_code = 3,
         year = 4, 
         deaths = 6,
         births = 7) |> 
  mutate(race = "t")

im_t <- bind_rows(imr_t_1, imr_t_2, imr_t_3, imr_t_4)

outcome_b <- im_t |> 
  mutate(imr = 1000 * deaths / births) |> 
  select(-deaths, -births, -state_code) |>
  drop_na(year) |> 
  rename(imr_t = imr) |> 
  select(-race) |> 
  filter(year %in% c(1995:2018)) |> 
  arrange(state, year) |> 
  left_join(fips, by = c("state" = "state_name")) |> 
  select(state = state.y,
         year,
         imr_t)

outcome <- left_join(outcome_a, outcome_b)
```

```{r}
save(bias, outcome, im, file = here("data", "visdata.Rdata"))
```

```{r}
complete_states <- outcome |> 
  group_by(state) |> 
  summarize(n = n()) |> 
  filter(n == 24) |> 
  pull(state)

outcome <- outcome |> 
  filter(state %in% complete_states) |> 
  ungroup() |> 
  filter(!(state %in% c("NE", "LA", "MA", "NV", "MD")))

#in addition to IMR, Nebraska was excluded because it has a unicameral legislature
#LA, MA, NV, MD had high missingness on gerrymandering data
#696 possible state-years (29 states, 24 years each)

save(outcome, file = here("data", "outcome.Rdata"))

```

going to end up with 4 different datasets

```{r}
lag0 <- left_join(outcome, preds, 
                   by = c("state" = "state", "year" = "year")) |> 
  left_join(covs,
            by = c("state" = "state", "year" = "year"))

lag1 <- left_join(outcome, preds, 
                   by = c("state" = "state", "year" = "lag1year")) |> 
  left_join(covs,
            by = c("state" = "state", "year" = "year"))

lag2 <- left_join(outcome, preds,
                   by = c("state" = "state", "year" = "lag2year")) |> 
  left_join(covs,
            by = c("state" = "state", "year" = "year"))

lag3 <- left_join(outcome, preds, 
                   by = c("state" = "state", "year" = "lag3year")) |> 
  left_join(covs,
            by = c("state" = "state", "year" = "year"))
```

missing data analysis

```{r}
lag0 |> 
  mutate(complete = complete.cases(lag0)) |> 
  filter(!complete) |> 
  nrow()

lag1 |> 
  mutate(complete = complete.cases(lag1)) |> 
  filter(!complete) |> 
  nrow()

lag2 |> 
  mutate(complete = complete.cases(lag2)) |> 
  filter(!complete) |> 
  nrow()

lag3 |> 
  mutate(complete = complete.cases(lag3)) |> 
  filter(!complete)
```

```{r}
lag0 |> 
  drop_na() |> 
  group_by(state) |> 
  summarize(n = n()) |> 
  arrange(n)

lag1 |> 
  drop_na() |> 
  group_by(state) |> 
  summarize(n = n()) |> 
  arrange(n)

lag2 |> 
  drop_na() |> 
  group_by(state) |> 
  summarize(n = n()) |> 
  arrange(n)

lag3 |> 
  drop_na() |> 
  group_by(state) |> 
  summarize(n = n()) |> 
  arrange(n)
```

```{r}
save(lag0, lag1, lag2, lag3, file = here("data", "df_aug2026.Rdata"))
```

# new outcomes

```{r}
### state policy and politics database
# https://asi.syr.edu/state-policy-politics-database-sppd/

### all cause working age mortality from CDC Wonder
load(here("data", "preds_covs.Rdata"))
sppd <- read_dta(here("data", "SPPD_V1_3.1.dta"))

sppd <- sppd |> 
  select(state, state_name, year, 
         sumopioid,
         tobaccotax_2023,
         firearms_total,
         medicaidexp,
         medgen_v1,
         medgen_v2,
         SNAP_2023,
         TANF2023) |> 
  filter(year %in% c(1997:2018))  |> 
  zap_formats() |>
  zap_labels() |> 
  zap_label()

acm <- readxl::read_xls(here("data", "workingagemortality.xls"))


acm <- acm |> 
  janitor::clean_names() |> 
  select(state_name = state,
         year,
         wam = crude_rate) |> 
  drop_na(year)

new_out <- left_join(sppd, acm, by = c("state_name", "year"))

df_nopreds <- left_join(new_out, covs, by = c("state", "year"))

newout_lag0 <- left_join(df_nopreds, preds, by = c("state" = "state", "year" = "year"))
newout_lag1 <- left_join(df_nopreds, preds, by = c("state" = "state", "year" = "lag1year"))

newout_lag2 <- left_join(df_nopreds, preds, by = c("state" = "state", "year" = "lag2year"))

newout_lag3 <- left_join(df_nopreds, preds, by = c("state" = "state", "year" = "lag3year"))

preds
save(newout_lag0, newout_lag1, newout_lag2, newout_lag3,
     file = here("data", "newout.Rdata"))
```
